---
title: "Resistência à penetração, conteúdo de água, porosidade e densidade de latossolos com diferente mineralogia"
author: >
  [Walmes Marques Zeviani](http://lattes.cnpq.br/4410617539281650) &
  [Carla Eloize Carducci](http://lattes.cnpq.br/3585988593213083)
date: '`r format(Sys.Date(), format = "%d de %B de %Y")`'
vignette: >
  %\VignetteIndexEntry{Resistência à penetração, conteúdo de água, porosidade e densidade de latossolos com diferente mineralogia}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
source("config/setup.R")
```

## Exploratory data analysis

```{r, eval = FALSE, include = FALSE}
devtools::load_all()
```
```{r, results = "hide", message = FALSE, error = FALSE, warning = FALSE}
#-----------------------------------------------------------------------
# Load packages.

rm(list = objects())

library(EACS)
library(lmerTest)
library(lme4)
library(car)
library(emmeans)
library(tidyverse)

# To compare curves with Wald F test statistic.
compare_curves <- function(index,
                           levels = NULL,
                           L = L,
                           ords = list(ord0, ord1)) {
    J <- L[index, , drop = FALSE]
    K <- wzRfun::apc(J, lev = levels)
    f_test <- apply(K[, , drop = FALSE],
                    MARGIN = 1,
                    FUN = function(Ki) {
                        U <- lapply(ords,
                                    function(ord) {
                                        Ki %*% diag(ord)
                                    })
                        U <- do.call(rbind, U)
                        f_test <-
                            linearHypothesis(model = m1,
                                             hypothesis.matrix = U,
                                             test = "F")
                        tail(f_test[, c("F", "Pr(>F)")], n = 1)
                    })
    f_test <- do.call(what = rbind, args = f_test)
    cbind(contrast = rownames(f_test), f_test)
}

# Text to be used as axis label.
axis_text <- list(
    tensao = expression(Log[10] ~ (Psi * "m, kPa") - "Matric tension"),
    thetai = expression("Water content" ~ (cm^{3} ~ cm^{-3})),
    pr = expression("Penetration resistance" ~ (MPa)),
    ds = expression("Bulk density" ~ (g ~ cm^{-3})),
    macro = expression("Macroporosity" ~ (cm^3 ~ cm^{-3})),
    micro = expression("Microporosity" ~ (cm^3 ~ cm^{-3})),
    total = expression("Total porosity" ~ (cm^3 ~ cm^{-3})),
    posi = "Position",
    solo = "Soil",
    prof = "Depth (cm)")
```

## Resistance to penetration and water content

```{r, include = FALSE}
txt <- "
solo	posi	pr	tensao	thetai
KH	interrrow	0,183	6	0,413140462349601
KH	interrrow	0,430	6	0,402379535990482
KH	interrrow	0,633	6	0,413611940298507
KH	interrrow	1,44634482758621	33	0,39975898334794
KH	interrrow	0,524053050397878	33	0,401574803149606
KH	interrrow	0,517814323607427	33	0,403754661180404
KH	interrrow	0,913583554376658	100	0,367959207873829
KH	interrrow	0,851456233421751	100	0,34668071654373
KH	interrrow	0,793877984084881	100	0,376058594644083
KH	interrrow	1,17807957559682	1500	0,340524703688907
KH	interrrow	0,777241379310345	1500	0,320927062031356
KH	interrrow	1,18028912466844	1500	0,331813701330705
KH	interrrow	1,06071352785146	3000	0,12070643860381
KH	interrrow	2,01484880636605	3000	0,116750613457316
KH	interrrow	2,001	3000	0,123170998034505
KH	row	0,438790450928382	6	0,40567612687813
KH	row	0,154928381962865	6	0,385090609555189
KH	row	0,183002652519894	6	0,37938224944611
KH	row	0,266185676392573	33	0,38792505204719
KH	row	0,48142175066313	33	0,365764631843927
KH	row	0,519893899204244	33	0,366102070187092
KH	row	0,398238726790451	100	0,331074977416441
KH	row	0,741368700265252	100	0,323778764634639
KH	row	0,284901856763926	100	0,327021026511689
KH	row	0,518854111405836	1500	0,323902737421624
KH	row	0,830010610079576	1500	0,277932607043324
KH	row	0,831960212201591	1500	0,30627661035572
KH	row	0,935809018567639	3000	0,113823973956517
KH	row	0,950366047745358	3000	0,129732868757259
KH	row	0,964923076923077	3000	0,149307059333045
GA	interrrow	0,265145888594164	6	0,506420909502946
GA	interrrow	0,166366047745358	6	0,480743293870823
GA	interrrow	0,549007957559682	6	0,464537642683083
GA	interrrow	0,438790450928382	33	0,437879533891351
GA	interrrow	0,608405835543767	33	0,402911960810995
GA	interrrow	0,270344827586207	33	0,380581728799672
GA	interrrow	0,62595225464191	100	0,407889033376679
GA	interrrow	0,551087533156499	100	0,412394291754757
GA	interrrow	0,794527851458886	100	0,349671175083757
GA	interrrow	1,16664190981432	1500	0,310809009729442
GA	interrrow	1,19159681697613	1500	0,308911973756151
GA	interrrow	1,51809018567639	1500	0,303158418303904
GA	interrrow	2,27180636604775	3000	0,084278992526986
GA	interrrow	2,20279045092838	3000	0,111649279682066
GA	interrrow	2,43362334217507	3000	0,125170574370426
GA	row	0,187161803713528	6	0,434439409443064
GA	row	0,581111405835544	6	0,455758055995774
GA	row	0,272424403183024	6	0,450909090909091
GA	row	0,308816976127321	33	0,372555107109593
GA	row	0,32337400530504	33	0,368315699170718
GA	row	0,456466843501326	33	0,36293827875978
GA	row	0,242270557029178	100	0,385232002466471
GA	row	0,251628647214854	100	0,372961755114142
GA	row	0,46062599469496	100	0,376149425287356
GA	row	1,18535809018568	1500	0,3105249745158
GA	row	1,61908	1500	0,294623655913979
GA	row	2,05280106100796	1500	0,31086189516129
GA	row	1,45986206896552	3000	0,422524449877751
GA	row	1,45986206896552	3000	0,107049228080396
GA	row	0,671053050397878	3000	0,200737693289828"
da <- read.table(textConnection(txt),
                 header = TRUE,
                 sep = "\t",
                 dec = ",")
str(da)

# Recode factor levels.
levels(da$posi) <- c("Interrow", "Row")
s <- c(KH = "Kaolinitic Haplustox", GA = "Gibbsitic Acrustox")
levels(da$solo) <- names(s)

# Recode levels.
da$trt <- with(da, interaction(solo, posi, sep = "·"))
da$logtensao <- log10(da$tensao)
da$rept <- 1:3

da <- da %>%
    mutate(ue1 = interaction(solo, rept),
           ue2 = interaction(solo, rept, posi))
str(da)
```

```{r}
# Data structure.
str(da)

# Cell frequency.
ftable(xtabs(~solo + posi + tensao, data = da))
```

### Penetration resistance

Linear mixed effects model were used to model soil resistance to
penetration as function of experimental factors. As suggested by
exploratory data analysis, the effect of matric tension can be properly
described by a second order degree polynomial.

```{r}
# Differnce in position mean profile conditional to other factors.
ggplot(data = da,
       mapping = aes(x = tensao, y = pr, color = posi)) +
    facet_grid(facets = ~solo) +
    geom_point() +
    stat_summary(geom = "line", fun.y = "mean") +
    scale_x_log10() +
    labs(x = axis_text$tensao,
         y = axis_text$pr,
         color = axis_text$posi)

# Differnce in position mean profile conditional to other factors.
ggplot(data = filter(da, tensao < 3000),
       mapping = aes(x = tensao, y = pr, color = posi)) +
    facet_grid(facets = ~solo) +
    geom_point() +
    geom_smooth(method = "lm",
                formula = y ~ poly(x, degree = 2),
                se = FALSE) +
    scale_x_log10() +
    labs(x = axis_text$tensao,
         y = axis_text$pr,
         color = axis_text$posi)

# Differnce in soil mean profile conditional to other factors.
ggplot(data = da,
       mapping = aes(x = tensao, y = pr, color = solo)) +
    facet_grid(facets = ~posi) +
    geom_point() +
    stat_summary(geom = "line", fun.y = "mean") +
    scale_x_log10() +
    labs(x = axis_text$tensao,
         y = axis_text$pr,
         color = axis_text$solo)
```

The statistical model is
$$
\begin{align*}
  y_{ijkl} &= \mu + S_i + P_j + \beta_1 T_k + \beta_2 T_k^2 + \\
   &\qquad (SP)_{ij} + (\beta_{1i} T_k + \beta_{2i} T_k^2) +
     (\beta_{1j} T_k + \beta_{2j} T_k^2) + \\
   &\qquad \beta_{1ij} T_k + \beta_{2ij} T_k^2 + \\
   &\qquad e_{il} + e_{ijl} + e_{ijkl}
\end{align*}
$$

  * $y_{ijkl}$ is response observed at the experimental condition $ijkl$.
  * $\mu$ is a constant that belongs to all observations.
  * $S_i$ is the effect of soil $i = 1, 2$.
  * $P_j$ is the effect of position $j = 1, 2$.
  * $\beta_1$ and $\beta_2$ are the first and second order effect of
    matric tension.
  * The following terms are two and three way interactions among soil,
    position and matric tension.
  * $e_{il}$ is the experimental error at experimental unit $il$,
    $e_{ijl}$ is the experimental error at experimental unit $ijl$ and
    $e_{ijkl}$ is the experimental error at sample unit $ijkl$.

This model were estimated by restricted maximum likelihood.

```{r}
# Random intercept, slope and curvature for random effects.
# Interaction until the 4th degree.
m0 <- lmer(pr ~ solo * posi * (logtensao + I(logtensao^2)) +
               (1 | ue1) + (1 | ue2),
           data = subset(da, tensao < 3000))

# Variance of the random effects.
VarCorr(m0)

# Test for the fixed effects terms.
anova(m0)

# Fitting a smaller model with relevant terms.
m1 <- update(m0,
             . ~ (posi + solo) *
                 (logtensao + I(logtensao^2)) +
                 # posi:solo +
                 (1 | ue1) + (1 | ue2))
anova(m0, m1)

# Test for the fixed effects terms.
anova(m1)
```

All three way interactions were dropped beacause were not relevant to
the model. Also, the two way interaction between soil and position were
not relevant and, then, dropped.

```{r}
# Estimated marginal means with covariate set at 1. The corresponding
# matrix will be used to get que equation of each experimental
# condition.
emm <- emmeans(m1,
               specs = ~solo + posi,
               at = list(logtensao = 1))

# Get experimental points and matrix to calculate marginal means.
grid <- attr(emm, "grid")
L <- attr(emm, "linfct")
t(unname(L))

# Fixed effects estimates.
b <- fixef(m1)

# Zero order terms (intercept).
ord0 <- names(b) %>%
    grepl(pattern = "logtensao") %>% `!`()

# First order terms (slope).
ord1 <- names(b) %>%
    grepl(pattern = "logtensao$")

# Second order terms (curvatures).
ord2 <- names(b) %>%
    grepl(pattern = "I\\(logtensao\\^2\\)$")

# Checking: all row sum must be 1.
rowSums(cbind(ord0, ord1, ord2))

# Get the equation coefficients: b0 + b1 * x + b2 * x^2.
grid$b0 <- c(L %*% (b * ord0))
grid$b1 <- c(L %*% (b * ord1))
grid$b2 <- c(L %*% (b * ord2))

# Equations.
grid <- grid %>%
    mutate(label = sprintf("'%0.2f' %s '%0.3f' * x %s '%0.4f' * x^2",
                           b0,
                           ifelse(b1 >= 0, "+", "-"),
                           abs(b1),
                           ifelse(b2 >= 0, "+", "-"),
                           abs(b2)))

# Table of equation coefficients for each experimental point.
grid

# Checking: are the equation curves correct? Compare with `predict()`.
grid_curves <- grid %>%
    group_by(posi, solo) %>%
    do({
        logtensao <- seq(0.5, 3.5, length = 31)
        fit <- .$b0 + .$b1 * logtensao + .$b2 * logtensao^2
        data.frame(logtensao, fit)
    })
# grid_curves
```

```{r}
# Conditions to get estimated values.
pred <- with(da,
             expand.grid(solo = levels(solo),
                         posi = levels(posi),
                         logtensao = seq(0.5, 3.5, length = 31),
                         KEEP.OUT.ATTRS = FALSE))

# Model matrix for prediction.
X <- model.matrix(formula(terms(m1))[-2], data = pred)

# Checking: length and names.
stopifnot(ncol(X) == length(fixef(m1)))
stopifnot(all(colnames(X) == names(fixef(m1))))

# Predicted value.
pred$fit <- X %*% fixef(m1)

# Standard error of predicted values.
pred$se <- sqrt(diag(X %*% tcrossprod(vcov(m1), X)))

# Quantile of t distribution.
qtl <- qt(p = 0.975, df = df.residual(m1))

# Confidence limits for predicted value.
pred$lwr <- pred$fit - qtl * pred$se
pred$upr <- pred$fit + qtl * pred$se

# Adds the equation to annotate on graphics.
grid <- grid %>%
    mutate(xpos = 2.5,
           ypos = ifelse(solo == "GA", 2.10, 2.02) - 0.1,
           hjust = c(0),
           vjust = c(0))

# Fitted curves with confidence bands and observed data.
# Soil in evidence.
ggplot(data = filter(da, tensao < 3000),
       mapping = aes(x = tensao, y = pr, color = solo)) +
    facet_grid(facets = ~posi) +
    geom_point() +
    scale_x_log10() +
    # geom_line(data = pred,
    #           mapping = aes(x = 10^logtensao,
    #                         y = fit,
    #                         color = solo)) +
    geom_smooth(data = pred,
                mapping = aes(x = 10^logtensao,
                              y = fit,
                              ymin = lwr,
                              ymax = upr),
                stat = "identity",
                show.legend = FALSE) +
    geom_text(data = grid,
              mapping = aes(x = xpos,
                            y = ypos,
                            label = label,
                            hjust = hjust,
                            vjust = vjust),
              show.legend = FALSE,
              parse = TRUE) +
    # geom_line(data = grid_curves,
    #           mapping = aes(x = 10^logtensao,
    #                         y = fit,
    #                         group = solo),
    #           linetype = 2,
    #           color = "black") +
    labs(x = axis_text$tensao,
         y = axis_text$pr,
         color = axis_text$solo)

grid <- grid %>%
    mutate(xpos = 2.5,
           ypos = ifelse(posi == "Interrow", 2.10, 2.02) + 0.1)

# Fitted curves with confidence bands and observed data.
# Position in evidence.
ggplot(data = filter(da, tensao < 3000),
       mapping = aes(x = tensao, y = pr, color = posi)) +
    facet_grid(facets = ~solo) +
    geom_point() +
    scale_x_log10() +
    geom_line(data = pred,
              mapping = aes(x = 10^logtensao,
                            y = fit,
                            color = posi)) +
    geom_smooth(data = pred,
                mapping = aes(x = 10^logtensao,
                              y = fit,
                              ymin = lwr,
                              ymax = upr),
                stat = "identity",
                show.legend = FALSE) +
    geom_text(data = grid,
              mapping = aes(x = xpos,
                            y = ypos,
                            label = label,
                            hjust = hjust,
                            vjust = vjust),
              show.legend = FALSE,
              parse = TRUE) +
    # geom_line(data = grid_curves,
    #           mapping = aes(x = 10^logtensao,
    #                         y = fit,
    #                         group = posi),
    #           linetype = 2,
    #           color = "black") +
    labs(x = axis_text$tensao,
         y = axis_text$pr,
         color = axis_text$posi)
```

```{r}
# Tests of soil curves are equal in each posi:prof combination.
grid %>%
    mutate(row = 1:n()) %>%
    group_by(posi) %>%
    do({
        compare_curves(.$row, .$solo, L, list(ord0, ord1, ord2))
    }) %>%
    as.data.frame()

# Tests of posi curves are equal in each soil:prof combination.
grid %>%
    mutate(row = 1:n()) %>%
    group_by(solo) %>%
    do({
        compare_curves(.$row, .$posi, L, list(ord0, ord1, ord2))
    }) %>%
    as.data.frame()
```

According to Wald F test to compare curve parameters, there isn't
enought evidence to reject the null hypothesis of curve equality between
sampling position. The result of the hypothesis test were the same for
both soils because soil and position did not interact.

### Water content

```{r}
# Differnce in position mean profile conditional to other factors.
ggplot(data = da,
       mapping = aes(x = tensao, y = thetai, color = posi)) +
    facet_grid(facets = ~solo) +
    geom_point() +
    stat_summary(geom = "line", fun.y = "mean") +
    scale_x_log10() +
    labs(x = axis_text$tensao,
         y = axis_text$thetai,
         color = axis_text$posi)

# Differnce in position mean profile conditional to other factors.
ggplot(data = filter(da, tensao < 3000),
       mapping = aes(x = tensao, y = thetai, color = posi)) +
    facet_grid(facets = ~solo) +
    geom_point() +
    geom_smooth(method = "lm",
                formula = y ~ poly(x, degree = 2),
                se = FALSE) +
    scale_x_log10() +
    labs(x = axis_text$tensao,
         y = axis_text$thetai,
         color = axis_text$posi)

# Differnce in soil mean profile conditional to other factors.
ggplot(data = da,
       mapping = aes(x = tensao, y = thetai, color = solo)) +
    facet_grid(facets = ~posi) +
    geom_point() +
    stat_summary(geom = "line", fun.y = "mean") +
    scale_x_log10() +
    labs(x = axis_text$tensao,
         y = axis_text$thetai,
         color = axis_text$solo)
```

```{r}
# Random intercept, slope and curvature for random effects.
# Interaction until the 4th degree.
m0 <- lmer(thetai ~ solo * posi * (logtensao + I(logtensao^2)) +
               # (1 + logtensao + I(logtensao^2) | ue),
               (1 | ue1) + (1 | ue2),
           data = subset(da, tensao < 3000))

# Variance of the random effects.
VarCorr(m0)

# Test for the fixed effects terms.
anova(m0)

# Fitting a smaller model with relevant terms.
m1 <- update(m0, . ~ posi + solo * logtensao +
                     (1 | ue1) + (1 | ue2))
anova(m0, m1)

# Test for the fixed effects terms.
anova(m1)
```

```{r}
# Estimated marginal means with covariate set at 1. The corresponding
# matrix will be used to get que equation of each experimental
# condition.
emm <- emmeans(m1,
               specs = ~solo + posi,
               at = list(logtensao = 1))

# Get experimental points and matrix to calculate marginal means.
grid <- attr(emm, "grid")
L <- attr(emm, "linfct")
t(unname(L))

# Fixed effects estimates.
b <- fixef(m1)

# Zero order terms (intercept).
ord0 <- names(b) %>%
    # grep(pattern = "logtensao", value = TRUE, invert = TRUE)
    grepl(pattern = "logtensao") %>% `!`()

# First order terms (slope).
ord1 <- names(b) %>%
    # grep(pattern = "logtensao$", value = TRUE, invert = FALSE)
    grepl(pattern = "logtensao$")

# Checking: all row sum must be 1.
rowSums(cbind(ord0, ord1))

# Get the equation coefficients: b0 + b1 * x.
grid$b0 <- c(L %*% (b * ord0))
grid$b1 <- c(L %*% (b * ord1))

grid <- grid %>%
    mutate(label = sprintf("'%0.2f' %s '%0.3f' * x",
                           b0,
                           ifelse(b1 >= 0, "+", "-"),
                           abs(b1)))

# Table of equation coefficients for each experimental point.
grid

# Checking: are the equation curves correct? Compare with `predict()`.
grid_curves <- grid %>%
    group_by(posi, solo) %>%
    do({
        logtensao <- seq(0.5, 3.5, length = 31)
        fit <- .$b0 + .$b1 * logtensao
        data.frame(logtensao, fit)
    })
# grid_curves
```

```{r}
# Conditions to get estimated values.
pred <- with(da,
             expand.grid(solo = levels(solo),
                         posi = levels(posi),
                         logtensao = seq(0.5, 3.5, length = 31),
                         KEEP.OUT.ATTRS = FALSE))

# Model matrix for prediction.
X <- model.matrix(formula(terms(m1))[-2], data = pred)

# Checking: length and names.
stopifnot(ncol(X) == length(fixef(m1)))
stopifnot(all(colnames(X) == names(fixef(m1))))

# Predicted value.
pred$fit <- X %*% fixef(m1)

# Standard error of predicted values.
pred$se <- sqrt(diag(X %*% tcrossprod(vcov(m1), X)))

# Quantile of t distribution.
qtl <- qt(p = 0.975, df = df.residual(m1))

# Confidence limits for predicted value.
pred$lwr <- pred$fit - qtl * pred$se
pred$upr <- pred$fit + qtl * pred$se

# Adds the equation to annotate on graphics.
grid <- grid %>%
    mutate(xpos = 2.5,
           ypos = ifelse(solo == "GA", 0.29, 0.30) + 0,
           hjust = c(0),
           vjust = c(0))

# Fitted curves with confidence bands and observed data.
# Soil in evidence.
ggplot(data = filter(da, tensao < 3000),
       mapping = aes(x = tensao, y = thetai, color = solo)) +
    facet_grid(facets = ~posi) +
    geom_point() +
    scale_x_log10() +
    # geom_line(data = pred,
    #           mapping = aes(x = 10^logtensao,
    #                         y = fit,
    #                         color = solo)) +
    geom_smooth(data = pred,
                mapping = aes(x = 10^logtensao,
                              y = fit,
                              ymin = lwr,
                              ymax = upr),
                stat = "identity",
                show.legend = FALSE) +
    geom_text(data = grid,
              mapping = aes(x = xpos,
                            y = ypos,
                            label = label,
                            hjust = hjust,
                            vjust = vjust),
              show.legend = FALSE,
              parse = TRUE) +
    # geom_line(data = grid_curves,
    #           mapping = aes(x = 10^logtensao,
    #                         y = fit,
    #                         group = solo),
    #           linetype = 2,
    #           color = "black") +
    labs(x = axis_text$tensao,
         y = axis_text$thetai,
         color = axis_text$solo)

grid <- grid %>%
    mutate(xpos = 2.5,
           ypos = ifelse(posi == "Interrow", 0.29, 0.30) + 0)

# Fitted curves with confidence bands and observed data.
# Position in evidence.
ggplot(data = filter(da, tensao < 3000),
       mapping = aes(x = tensao, y = thetai, color = posi)) +
    facet_grid(facets = ~solo) +
    geom_point() +
    scale_x_log10() +
    geom_line(data = pred,
              mapping = aes(x = 10^logtensao,
                            y = fit,
                            color = posi)) +
    geom_smooth(data = pred,
                mapping = aes(x = 10^logtensao,
                              y = fit,
                              ymin = lwr,
                              ymax = upr),
                stat = "identity",
                show.legend = FALSE) +
    geom_text(data = grid,
              mapping = aes(x = xpos,
                            y = ypos,
                            label = label,
                            hjust = hjust,
                            vjust = vjust),
              show.legend = FALSE,
              parse = TRUE) +
    # geom_line(data = grid_curves,
    #           mapping = aes(x = 10^logtensao,
    #                         y = fit,
    #                         group = posi),
    #           linetype = 2,
    #           color = "black") +
    labs(x = axis_text$tensao,
         y = axis_text$thetai,
         color = axis_text$posi)
```

```{r}
# Tests of soil curves are equal in each posi:prof combination.
grid %>%
    mutate(row = 1:n()) %>%
    group_by(posi) %>%
    do({
        compare_curves(.$row, .$solo, L, list(ord0, ord1))
    }) %>%
    as.data.frame()

# Tests of posi curves are equal in each soil:prof combination.
grid %>%
    mutate(row = 1:n()) %>%
    group_by(solo) %>%
    do({
        compare_curves(.$row, .$posi, L, list(ord0, ord1))
    }) %>%
    as.data.frame()
```

## Porosity and bulk density

```{r, include = FALSE}
txt <- "
solo	prof	posi	rep	total	micro	macro	ds
GA	0-5	row	1	0,66	0,37	0,29	0,84
GA	0-5	row	2	0,63	0,43	0,19	0,82
GA	0-5	row	3	0,66	0,39	0,28	0,99
GA	0-5	row	4	0,68	0,37	0,31	0,84
GA	0-5	row	5	0,66	0,43	0,23	0,95
GA	0-5	row	6	0,66	0,39	0,28	0,99
GA	0-5	row	7	0,68	0,37	0,31	0,83
GA	0-5	row	8	0,67	0,43	0,24	0,92
GA	0-5	row	9	0,65	0,39	0,26	0,91
GA	0-5	row	10	0,61	0,37	0,24	0,86
GA	0-5	row	11	0,66	0,43	0,23	0,95
GA	0-5	row	12	0,61	0,39	0,22	0,86
GA	0-5	interrow	1	0,67	0,42	0,25	0,85
GA	0-5	interrow	2	0,68	0,39	0,29	0,99
GA	0-5	interrow	3	0,61	0,46	0,16	0,87
GA	0-5	interrow	4	0,70	0,42	0,27	0,8
GA	0-5	interrow	5	0,64	0,39	0,25	0,86
GA	0-5	interrow	6	0,64	0,46	0,18	0,89
GA	0-5	interrow	7	0,66	0,42	0,24	0,78
GA	0-5	interrow	8	0,63	0,39	0,24	0,91
GA	0-5	interrow	9	0,61	0,46	0,16	0,91
GA	0-5	interrow	10	0,63	0,42	0,21	0,89
GA	0-5	interrow	11	0,65	0,39	0,26	1,05
GA	0-5	interrow	12	0,60	0,46	0,14	0,98
GA	10-15	row	1	0,67	0,38	0,28	0,85
GA	10-15	row	2	0,61	0,43	0,18	0,93
GA	10-15	row	3	0,66	0,40	0,26	0,91
GA	10-15	row	4	0,68	0,38	0,29	0,81
GA	10-15	row	5	0,66	0,43	0,23	0,87
GA	10-15	row	6	0,66	0,40	0,26	0,86
GA	10-15	row	7	0,68	0,38	0,30	0,82
GA	10-15	row	8	0,66	0,43	0,23	0,88
GA	10-15	row	9	0,67	0,40	0,27	0,87
GA	10-15	row	10	0,67	0,38	0,29	0,92
GA	10-15	row	11	0,66	0,43	0,23	0,87
GA	10-15	row	12	0,68	0,40	0,28	0,37
GA	10-15	interrow	1	0,68	0,37	0,32	0,86
GA	10-15	interrow	2	0,63	0,40	0,23	0,94
GA	10-15	interrow	3	0,61	0,47	0,15	0,99
GA	10-15	interrow	4	0,65	0,37	0,29	0,93
GA	10-15	interrow	5	0,59	0,40	0,19	1,03
GA	10-15	interrow	6	0,62	0,47	0,15	0,91
GA	10-15	interrow	7	0,64	0,37	0,27	0,87
GA	10-15	interrow	8	0,60	0,40	0,20	0,89
GA	10-15	interrow	9	0,65	0,47	0,18	0,94
GA	10-15	interrow	10	0,64	0,37	0,27	0,81
GA	10-15	interrow	11	0,60	0,40	0,20	0,84
GA	10-15	interrow	12	0,64	0,47	0,17	0,89
GA	20-25	row	1	0,68	0,37	0,32	0,82
GA	20-25	row	2	0,66	0,39	0,27	0,87
GA	20-25	row	3	0,65	0,39	0,26	0,84
GA	20-25	row	4	0,64	0,37	0,27	0,88
GA	20-25	row	5	0,66	0,39	0,27	0,89
GA	20-25	row	6	0,63	0,39	0,24	0,84
GA	20-25	row	7	0,65	0,37	0,28	0,94
GA	20-25	row	8	0,65	0,39	0,26	0,9
GA	20-25	row	9	0,67	0,39	0,28	1
GA	20-25	row	10	0,62	0,37	0,25	0,93
GA	20-25	row	11	0,66	0,39	0,27	1
GA	20-25	row	12	0,67	0,39	0,29	0,93
GA	20-25	interrow	1	0,67	0,39	0,29	0,84
GA	20-25	interrow	2	0,64	0,42	0,23	0,95
GA	20-25	interrow	3	0,64	0,46	0,19	0,98
GA	20-25	interrow	4	0,67	0,39	0,28	0,99
GA	20-25	interrow	5	0,64	0,42	0,22	0,87
GA	20-25	interrow	6	0,64	0,46	0,19	0,99
GA	20-25	interrow	7	0,66	0,39	0,27	0,83
GA	20-25	interrow	8	0,65	0,42	0,24	0,86
GA	20-25	interrow	9	0,63	0,46	0,18	0,81
GA	20-25	interrow	10	0,67	0,39	0,29	0,96
GA	20-25	interrow	11	0,63	0,42	0,21	0,86
GA	20-25	interrow	12	0,62	0,46	0,16	0,83
KH	0-5	row	1	0,63	0,39	0,24	0,9
KH	0-5	row	2	0,65	0,35	0,30	0,99
KH	0-5	row	3	0,64	0,36	0,28	0,95
KH	0-5	row	4	0,66	0,39	0,27	0,87
KH	0-5	row	5	0,63	0,35	0,28	1,06
KH	0-5	row	6	0,57	0,36	0,21	0,91
KH	0-5	row	7	0,68	0,39	0,29	0,86
KH	0-5	row	8	0,64	0,35	0,29	0,92
KH	0-5	row	9	0,64	0,36	0,28	0,84
KH	0-5	row	10	0,69	0,39	0,30	0,82
KH	0-5	row	11	0,63	0,35	0,28	0,85
KH	0-5	row	12	0,65	0,36	0,29	0,87
KH	0-5	interrow	1	0,64	0,38	0,26	0,84
KH	0-5	interrow	2	0,59	0,42	0,16	0,86
KH	0-5	interrow	3	0,59	0,43	0,15	0,92
KH	0-5	interrow	4	0,55	0,38	0,17	0,89
KH	0-5	interrow	5	0,61	0,42	0,18	0,86
KH	0-5	interrow	6	0,61	0,43	0,18	0,83
KH	0-5	interrow	7	0,59	0,38	0,21	0,92
KH	0-5	interrow	8	0,63	0,42	0,21	1,05
KH	0-5	interrow	9	0,57	0,43	0,13	1,05
KH	0-5	interrow	10	0,64	0,38	0,25	1,12
KH	0-5	interrow	11	0,64	0,42	0,22	0,98
KH	0-5	interrow	12	0,61	0,43	0,17	1
KH	10-15	row	1	0,66	0,38	0,28	1,01
KH	10-15	row	2	0,67	0,35	0,32	0,98
KH	10-15	row	3	0,64	0,37	0,27	1,13
KH	10-15	row	4	0,64	0,38	0,27	0,97
KH	10-15	row	5	0,65	0,35	0,30	0,92
KH	10-15	row	6	0,62	0,37	0,25	0,95
KH	10-15	row	7	0,68	0,38	0,30	0,91
KH	10-15	row	8	0,62	0,35	0,27	0,87
KH	10-15	row	9	0,65	0,37	0,28	0,95
KH	10-15	row	10	0,64	0,38	0,26	0,99
KH	10-15	row	11	0,64	0,35	0,29	0,93
KH	10-15	row	12	0,65	0,37	0,27	1,15
KH	10-15	interrow	1	0,56	0,45	0,11	1,14
KH	10-15	interrow	2	0,62	0,38	0,23	1
KH	10-15	interrow	3	0,61	0,39	0,22	0,99
KH	10-15	interrow	4	0,57	0,45	0,12	1,1
KH	10-15	interrow	5	0,63	0,38	0,24	0,95
KH	10-15	interrow	6	0,60	0,39	0,20	1,03
KH	10-15	interrow	7	0,56	0,45	0,10	0,95
KH	10-15	interrow	8	0,62	0,38	0,23	1,04
KH	10-15	interrow	9	0,62	0,39	0,23	1,09
KH	10-15	interrow	10	0,57	0,45	0,12	0,9
KH	10-15	interrow	11	0,65	0,38	0,26	0,98
KH	10-15	interrow	12	0,60	0,39	0,21	1,13
KH	20-25	row	1	0,63	0,40	0,23	0,95
KH	20-25	row	2	0,65	0,38	0,27	0,92
KH	20-25	row	3	0,56	0,42	0,14	1,01
KH	20-25	row	4	0,64	0,40	0,25	0,95
KH	20-25	row	5	0,59	0,38	0,21	1,08
KH	20-25	row	6	0,63	0,42	0,21	0,97
KH	20-25	row	7	0,63	0,40	0,23	1,05
KH	20-25	row	8	0,64	0,38	0,26	0,94
KH	20-25	row	9	0,64	0,42	0,22	1,1
KH	20-25	row	10	0,63	0,40	0,24	1,13
KH	20-25	row	11	0,65	0,38	0,27	0,97
KH	20-25	row	12	0,66	0,42	0,24	0,97
KH	20-25	interrow	1	0,60	0,43	0,17	1,01
KH	20-25	interrow	2	0,59	0,42	0,17	1,03
KH	20-25	interrow	3	0,56	0,41	0,14	1,18
KH	20-25	interrow	4	0,63	0,43	0,20	0,84
KH	20-25	interrow	5	0,59	0,42	0,17	0,95
KH	20-25	interrow	6	0,57	0,41	0,16	0,95
KH	20-25	interrow	7	0,60	0,43	0,17	0,86
KH	20-25	interrow	8	0,60	0,42	0,18	1,02
KH	20-25	interrow	9	0,54	0,41	0,12	0,92
KH	20-25	interrow	10	0,60	0,43	0,17	0,98
KH	20-25	interrow	11	0,58	0,42	0,15	0,95
KH	20-25	interrow	12	0,56	0,41	0,15	0,95"
da <- read.table(textConnection(txt),
                 header = TRUE,
                 sep = "\t",
                 dec = ",")

# Recode factor levels.
levels(da$posi) <- c("Interrow", "Row")
s <- c(KH = "Kaolinitic Haplustox", GA = "Gibbsitic Acrustox")
levels(da$solo) <- names(s)

# Recode levels.
da$rept <- factor(rep(1:3, times = 4))
str(da)

da <- da %>%
    group_by(solo, posi, prof, rept) %>%
    summarise_at(vars(total, macro, micro, ds), "mean") %>%
    ungroup()

da$trt <- with(da, interaction(solo, posi, prof, sep = "·"))

da <- da %>%
    mutate(ue1 = interaction(solo, rept),
           ue2 = interaction(solo, rept, posi))
str(da)
```

```{r}
# Data structure.
str(da)

# ATTENTION: rep is pseudo replication or not?

# Frequency of cell combination.
ftable(xtabs(~solo + posi + prof, data = da))

db <- da %>%
    gather(key = "variable", value = "value", total, macro, micro, ds)

ggplot(data = db,
       mapping = aes(x = prof, y = value, color = posi, group = posi)) +
    facet_grid(facets = variable ~ solo, scale = "free_y") +
    geom_jitter(width = 0.1, height = 0) +
    stat_summary(geom = "line", fun.y = "mean")

# To keep results.
tb_results <- list()
```

### Soil density

```{r}
m0 <- lmer(ds ~ solo * posi * prof + (1 | ue1) + (1 | ue2),
           data = da)

VarCorr(m0)

summary(m0)
anova(m0)

m1 <- update(m0,
             . ~ (solo + posi + prof)^2 - solo:posi +
                 (1 | ue1) + (1 | ue2),
             data = da)
anova(m0, m1)

anova(m1)

emm <- emmeans(m1, ~prof | solo)
as.data.frame(emm)

results <- multcomp::cld(emm)
results

results <- results %>%
    as.data.frame() %>%
    mutate(cld = c("a", "a", "a", "b", "a", "a"))

ggplot(data = results,
       mapping = aes(x = prof,
                     y = emmean,
                     ymin = lower.CL,
                     ymax = upper.CL)) +
    facet_wrap(facets = ~solo) +
    geom_point() +
    geom_errorbar(width = 0.1) +
    geom_text(mapping = aes(label = sprintf("%0.2f %s",
                                            emmean,
                                            cld)),
              angle = 90, vjust = -0.5) +
    labs(x = axis_text$prof,
         y = axis_text$ds)

tb_results$ds <- results

```

### Total porosity

```{r}
m0 <- lmer(total ~ solo * posi * prof + (1 | ue1) + (1 | ue2),
           data = da)

VarCorr(m0)
anova(m0)

m1 <- update(m0,
             . ~ solo + posi +
                 (1 | ue1) + (1 | ue2),
             data = da)
anova(m0, m1)

anova(m1)

emm <- emmeans(m1, ~posi | solo)
as.data.frame(emm)

results <- multcomp::cld(emm)
results

results <- results %>%
    as.data.frame() %>%
    mutate(cld = c("b", "a", "b", "a"))

ggplot(data = results,
       mapping = aes(x = posi,
                     y = emmean,
                     ymin = lower.CL,
                     ymax = upper.CL)) +
    facet_wrap(facets = ~solo) +
    geom_point() +
    geom_errorbar(width = 0.1) +
    geom_text(mapping = aes(label = sprintf("%0.2f %s",
                                            emmean,
                                            cld)),
              angle = 90, vjust = -0.5) +
    labs(x = axis_text$posi,
         y = axis_text$total)

tb_results$total <- results
```

### Macro porosity

```{r}
m0 <- lmer(macro ~ solo * posi * prof + (1 | ue1) + (1 | ue2),
           data = da)

VarCorr(m0)
anova(m0)

m1 <- update(m0,
             . ~ posi +
                 (1 | ue1) + (1 | ue2),
             data = da)
anova(m0, m1)

anova(m1)

emm <- emmeans(m1, ~posi)
as.data.frame(emm)

results <- multcomp::cld(emm)
results

results <- results %>%
    as.data.frame() %>%
    mutate(cld = c("b", "a"))

ggplot(data = results,
       mapping = aes(x = posi,
                     y = emmean,
                     ymin = lower.CL,
                     ymax = upper.CL)) +
    geom_point() +
    geom_errorbar(width = 0.1) +
    geom_text(mapping = aes(label = sprintf("%0.2f %s",
                                            emmean,
                                            cld)),
              angle = 90, vjust = -0.5) +
    labs(x = axis_text$posi,
         y = axis_text$macro)

tb_results$macro <- results
```

### Micro porosity

```{r}
m0 <- lmer(micro ~ solo * posi * prof + (1 | ue1) + (1 | ue2),
           data = da)

VarCorr(m0)
anova(m0)

m1 <- update(m0,
             . ~ posi +
                 (1 | ue1) + (1 | ue2),
             data = da)
anova(m0, m1)

anova(m1)

emm <- emmeans(m1, ~posi)
as.data.frame(emm)

results <- multcomp::cld(emm)
results

results <- results %>%
    as.data.frame() %>%
    mutate(cld = c("b", "a"))

ggplot(data = results,
       mapping = aes(x = posi,
                     y = emmean,
                     ymin = lower.CL,
                     ymax = upper.CL)) +
    geom_point() +
    geom_errorbar(width = 0.1) +
    geom_text(mapping = aes(label = sprintf("%0.2f %s",
                                            emmean,
                                            cld)),
              angle = 90, vjust = -0.5) +
    labs(x = axis_text$posi,
         y = axis_text$micro)

tb_results$micro <- results
```

```{r}
tb <- tb_results %>%
    bind_rows()
```

## Session information

```{r, echo = FALSE, results = "hold"}
cat(format(Sys.time(),
           format = "Atualizado em %d de %B de %Y.\n\n"))
sessionInfo()
```
